---
title: "Population Scale Analysis"
author: "Dora Deng (A17445600)"
format: pdf
---
Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

read the data: 
```{r}
expr <- read.table("https://bioboot.github.io/bimm143_W25/class-material/rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
Figure out the sample size of each genotype. 
```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Use boxplot() to extract median expression levels

```{r}
boxplot <- boxplot(exp ~ geno, expr) 
boxplot_medians <- boxplot$stats[3, ]  
names(boxplot_medians) <- levels(as.factor(expr$geno))
boxplot_medians
```

```{r}
summary(expr$exp)
```

Q14.Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.

Lets make a boxplot.

```{r}
ggplot(expr) + aes(x = geno, y = exp, fill = geno) + 
  geom_boxplot(notch = TRUE)
```

Since there are difference in the medium for each genotype, genotypes do affect the gene expression. SNP does affect the expression of ORMDL3, with A/A has a higher medium value and G/G has a comparatively lower medium value. 
