---
title: "Comparative Structure analysis"
author: "Dora Deng (A17445600)"
format: html
---

Here we run through a complete "pipeline" of structure analysis that begins with a single sequence identifer and ends in a PCA analysis. 

```{r}
library(bio3d)
```

## STEP 1. retrive a sequence for the protein we are interested in. We will take ADK "1ake_A"

```{r}
id <- "1ake_A" 
aa <- get.seq(id)
aa
```

## STEP 2. BLAST search 
Run a BLAST search of the PDB for all related sequences to our input `aa`
```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```

## STEP 3. Download all structures
These are our "top hits" i.e. all the structures in the PDB database related to our input sequence. 

```{r}
hits$pdb.id
```

```{r}
files <- get.pdb(hits$pdb.id, path = "pdbs", split=TRUE, gzip=TRUE)
```

## STEP 4. Align and superpose 

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")

```

```{r}
pdbs
```

## STEP 5. PCA

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships. 

```{r}
pc <- pca(pdbs)
plot(pc)
```

Let's make a trajectory (or movie) of the main confrontational changes captured by PC1. We will use the `mktrj()` function for this 

```{r}
mktrj(pc, file = "pca_result.pdb")
```

Back of the envelope comparison of the PDB size to UniProt

```{r}
uniprot <- 253206171
pdb <- 231029
pdb/uniprot * 100
```




